The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-10-13 adaptation/article/article.tex
Update on Overleaf.

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}
% \documentclass{article, a4paper}
\usepackage[font=small]{caption}
\usepackage{lscape}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{soul}
\usepackage{todonotes}

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we use a computational mechanistic model to explain the behavioral phenomena effected by imperfect contextual information. Specifically, we present a hierarchical computational model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect adaptation and action selection. We show how the proposed model provides a unifying explanation for four different experimentally established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of previously-learned adaptations on switching between tasks, in addition to  (iii) the well-studied savings, de-adaptation and spontaneous recovery.

@@

By introducing blocks of trials in which a manipulation is present (e.g. a mechanical arm exerts a force), experimenters are able to observe the dynamics of adaptation through the lens of motor error. Across many different motor adaptation experiments, now well-established phenomena have been observed, for example: (i) savings (the ability to recall previously-learned skills), (ii) quick de-adaptation (the ability to return to unmodified dynamics), (iii) anterograde interference (the interference in motor learning between opposing manipulations in dynamics) and (iv) spontaneous recovery (during trials where errors are forced to be zero, the display of behavior consistent with the previously-learned dynamics).

@@

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. Intuitively, in mechanical arm experiments, one can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference is probabilistic and takes the form of an evidence-accumulating process. In this work, we propose that the behavioral consequences of this slow process of evidence accumulation can be observed in behavior and can be used to explain the phenomena outlined above.

@@

Central to the model is the concept of context. In our model, contexts are defined in terms of the task to be performed (see \fref{fig:model}A), the variables of the environment that are relevant to perform the task, the forward models used by the decision-making agent to perform the task, and the update mechanisms necessary to adapt these forward models to the variables of the environment. Together, these elements allow the agent to make predictions on future observations when this context is active, and these predictions are used to infer the context. For example, when lifting an object of unknown weight, an agent might have learned one context for heavy objects and one for light objects. When observing an object to be lifted, the agent can use its size and texture to estimate the weight of the object, which in turn allows the agent to infer the appropriate context and, with it, decide how to lift the object. These contextual cues (size and texture) do not necessarily identify the context with certainty (e.g. it could be a movie prop that looks like a rock) and integration over time might be necessary.

@@

It is widely accepted that in order to adapt motor commands to a novel environment, we create and update internal forward models of the outcomes of control signals \citep{Wolpert_Multiple_1998}. Here, we use this principle and chose to make use of an exact Bayesian learner. As has been discussed before \cite{Oh_Minimizing_2019}, a Bayesian learner has the advantage of not only fitting experimental data on motor adaptation, but also eliminating the need for explicit multiple time scales of learning as the learner can deal with adaptation after any number of trials. As we will show below this feature will be useful to explain specific key phenomena parsimoniously.

To describe this component of the model, we make use of the example of reaching movements. We define a forward model as follows:
\begin{equation}
p(\vartheta | s, c) = f(s, c; \phi) \label{eqn:forward-model}
\end{equation}
where $\vartheta$ are the outcomes of an action (e.g. the final position of the hand in a reaching movement), $s$ is the current state (the current position of the hand), $c$ is a motor command and $\phi$ are the parameters of the forward model $f(\cdot)$. We assume that during learning the parameters $\phi$ have been fine tuned to produce an accurate forward model $f$ to guide our movements and it is these parameters which are updated during motor adaptation.

@@

\subsection{From observation to action selection}
Each trial starts with a new observation $\vartheta_t$, which comprises contextual sensory information (when available), as well as the outcome of the previously-selected action.

@@

\section{Experimental results}
In this section, we present experimentally-observed phenomena in four sections, and show that context inference, as done in our model, provides a unifying explanation for all of them. In the first section, we discuss switches between contexts, and how slow context inference affects these switches. In the second section, we focus on the effects of volatility in the environment on motor adaptation. In the third section, we discuss the effects of slow context inference on action selection. Finally, in the fourth section we discuss context inference during so-called error-clamp trials, and its effect on behavior.

@@

\subsection{Savings and slow/fast switching}
The term 'savings' refer to the ability of humans and other animals to remember a previously-learned adaptation and apply it without having to re-learn it. Savings is almost universally observed in human participants \cite{Brashers-Krug_Consolidation_1996,Shadmehr_Functional_1997,Medina_Mechanism_2001,Smith_Interacting_2006,Zarahn_Explaining_2008}. In an O-A-O-A experiment, for example, savings would express themselves in the second A block in the form of a much higher adaptation rate higher than that observed during the first A block.

@@

To show how adaptation changes depending on the amount of contextual information, we selected three representative experiments from two studies \cite{Kim_Neural_2015,Oh_Minimizing_2019} which differ in the amount of contextual information. \cite{Kim_Neural_2015} performed a cued-context visuomotor rotation experiment with three contexts with different rotation: no rotation, 40 degrees and -40 degrees. Participants performed shooting movements in blocks of trials with the same rotation. Importantly, the identity of the current context was displayed as a visual cue (colored light), making this a cued-context experiment. Consistent with our model, the authors found that switching was immediate and accurate. In \fref{fig:oh-2019}A we show the results of simulations with our model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts, i.e. when the solid black line (representing the true context) switches between 40, 0 and -40; in both panels (left and right), the blue line, representing the agent's adaptation, quickly follows the switches.

@@

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (true rotation). In the second experiment (\fref{fig:oh-2019}C), switching happens more slowly, with adaptation (blue line) lagging behind the switches in the real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained by our model in terms of the size of the adaptation: as the adaptation is smaller (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\subsection{Uncertainty over contexts affects action selection}
As with learning, our model selects actions (motor commands) based on context inference. If the identity of the current context is known, the forward model for this context is used to select the current action. However, if uncertainty over the context exists, the selected action is influenced by all the possible contexts, with a weight directly related to how likely each one of those contexts is (see \eref{eq:action-selection}).

@@

\subsection{Action selection in error-clamp blocks}
During error-clamp blocks at the end of block sequences, participants' behavior can be divided in two phases: (1) Spontaneous recovery, during which participants' behavior is consistent with a previously-encountered context (e.g. in an O-A-E experiment, consistent with A); this phase, when present, is seen during the early trials of the E block. (2) A slow return to baseline (i.e. no observable adaptation, consistent with the baseline context), which can last as long as hundreds of trials \citep{Brennan_Decay_2015}. However, the direction of the spontaneous recovery, its duration, the delay before it is observed, the speed of the return to baseline and the final asymptote of the adaptation seem to vary greatly depending on the experiment \citep{Brennan_Decay_2015,Vaswani_Decay_2013,Smith_Interacting_2006,Shmuelof_Overcoming_2012}.

@@

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). The start and speed of the drop dependd on the run (i.e. the participant), reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

\subsubsection{Volatility has no effect on learning}
The volatility of the environment (i.e. unpredictable switching between contexts) lowers the measured learning rate. Our model predicts that this effect would be greatly reduced if reliable contextual cues were introduced: if a participant can infer the context of the current trial before any decision or observation has occurred, the learning rate would not be affected by the volatility of the environment. This would manifest in \fref{fig:herzfeld-2014} in the form of no discernible differences between the two groups of participants.

@@

\subsubsection{Multi-source integration}
The model also predicts an effect reminiscent of multisensory integration \citep[e.g.][]{Ernst_Humans_2002}: in order to integrate contextual information from conflicting sources (e.g. probabilistic visual cues and noisy endpoint feedback), the weight placed on a source increases with its precision (the inverse of its variance). Such integration would manifest itself in experiments in which observations are noisy, as in the experiments by \cite{Kording_Bayesian_2004}, in which the position of the finger was obscured and instead participants are shown a blurry cursor which was some times shifted from its real position. If the added observation noise gives evidence for a particular context (the true underlying context or another one) and a visual cue gave partial information for another context, the participants' behavior would be more consistent with the most reliable source of contextual information.

@@

Before proceeding with the model description, let us introduce nomenclature:
\begin{itemize}
\item $t$: Time at the moment an observation is made, before inference is done.
\item $\vartheta_t$: Observation. Note that this is assumed to be a noisy
observation of the generalized state (see below) given by $\vartheta_t = h(z_t)$.
\item $s_t$: Hidden state, e.g. the position of the hand in reaching
experiments. It does not depend on the context.
\item $c_t$: Motor command (action) taken after observing $\vartheta_t$ and inferring the context.
\item $\zeta_t$: Context inferred after observing $\vartheta_t$.
\item $\omega_t$: Contextual information. This term includes, e.g. visual cues,
proprioceptive feedback, performance feedback.
\item $z_t$: Generalized state, $z_t = \{s_t, \zeta_t, \omega_t\}$. Observed noisily
as $\vartheta_t$.
\end{itemize}

Context inference can work through context-specific (possibly unreliable) sensory cues,
or through the dynamics of the system (e.g. expected outputs of motor
commands), using prediction error as the force behind inference. In this case,
the generative model can be written as follows:
\[
p(z_t, z_{t-1}, c_{t-1}, \vartheta_t, \vartheta_{t-1}) = p(\vartheta_t | z_t)p(z_t|\vartheta_{t-1}, c_{t-1})
\]
and inferring the current (hidden) generalized state is done with Bayes' theorem:
\[
  q(z_t | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | z_t)p(z_t|\vartheta_{t-1}, c_{t-1})
\]

Additionally, after having inferred the current state, motor adaptation is given by
\[
q(\gamma | z_t, z_{t-1}, c_{t-1}) \propto p(z_t | z_{t-1}, c_{t-1}, \phi)p(z_{t-1})p(\phi)
\]

where $\phi$ are the parameters of the internal model for the dynamics,
i.e. the function $f: (z_t, c_t) \rightarrow z_{t+1}$, which determines the beliefs of
the agent regarding how the system evolves after having taken action $c_t$.

Having inferred the generalized state (including context), a motor command is
sampled from a distribution:
\begin{equation}
p(c_t | z_t) = \displaystyle\sum_{m \in M}q(\zeta_t)p(c_t | z_t, m)
\end{equation}
where $M$ is the set of all relevant forward models (see below), and
$p(a_t | ... m)$ is the probability distribution over actions prescribed by the
forward model $m$.

@@

\subsubsection{Mathematical implementation}
The generalized state is given by:
\begin{equation}
z_t = \{s, \vartheta_t, \phi_t, \omega_t\}
\end{equation}
where $s$ is, e.g. in a reaching task, the current position of the hand in
Cartesian coordinates, where the origin is the starting point. We assume that
motor commands are issued every $\Delta t$ for simplicity. At the beginning of
each time interval, the context is inferred combining the contextual
information and the prediction error of the outcome of the previous motor
command. This yields:
\begin{align}
  \begin{split}
  q(\zeta_t) &= q(\zeta_t | \omega_t)q(\zeta_t | s_t, s_{t-1}, c_{t-1}) \\ \label{eqn:estimated-context}
  q(\zeta_t | \omega_t) &\propto p(\omega_t | \zeta_t)p(\zeta_t) \\
  q(\zeta_t | s_t, s_{t-1}, c_{t-1}) &\propto p(s_t | \zeta_t, s_{t-1}, c_{t-1})p(x_t)p(s_{t-1})p(\zeta_t)
  \end{split}
\end{align}
The terms $p(\zeta_t)$ refer to the prior probability of the context at the
beginning of the trial: at the first trial, this refers to prior beliefs over
which contexts are more common/likely, which we call $p_0(\zeta_t)$. At each
subsequent trial, it incorporates the belief that has so far been accumulated,
given previous trials. Note that certain events, such as the start of a new
block of trials (see main text) can return an agent to a state in which
$p_0(\zeta_t)$ becomes relevant.

We do not directly model hand movements or eye position. Instead, we model
systems in the motor error space: the position $s_t = 0$ represents a trial in
which no motor error occurred. We make use of the fact that in most
experiments, motor adaptation needs to happen in one of two directions
(e.g. clockwise vs. counter-clockwise) to keep the error space
two-dimensional. With this, $s_t > 0$ represents motor error in one direction
(e.g. clockwise) and $s_t < 0$, error in the other direction. We made this
choice because, in this work modeled different experiments with
different state spaces, but with the commonality of (at most)
two-dimensional error spaces. In order to model a specific experiment in its
own physical space (e.g. hand positions in a reaching task), only the dynamical
system would need to be adapted.

We assume that the decision-making agent ha a baseline internal forward models that
map motor commands to outcomes:
\begin{equation}
p(s_{t+1} | s_t, c_t, m) = f_O(s_t, c_t)
\end{equation}
where $c_t$ are motor commands (actions). We further assume that participants
have learned $f_O$ throughout their lives and use it to decide which motor
command to issue under normal circumstances (i.e. no adaptation). When a
perturbation is introduced, to which the participant must adapt, we assume they
learn a new forward model $f_m^{\phi}$, where $\phi$ represents the parameters
of the function that participants must learn. For example, in a saccade task,
$\phi$ represents the new gain. Note that $\phi$ depends on the context and
participants may need to use and update multiple forward adapted models
$f_m^{\phi}$ throughout one experiment.

@@

To complete the action-selection picture, we return to context inference. As
discussed above, action selection is made via a weighted sum of the different
forward models:
\begin{equation}
p(c_t | ...) \propto \displaystyle \sum_{m \in M} q(\zeta_i)p(c_t | m)
\end{equation}
From this distribution, an action is sampled at each trial.

@@

In the following equations, we drop the dependence on the context and
observations for notational simplicity:
\begin{equation}
q(\phi | s_{1:t}) \propto p(s_{1:t} | \phi)p(\phi)
\end{equation}
where $s_{1:t}$ are the observed states up until time $t$. The likelihood is a
Gaussian:
\begin{align}
  p(s_t | \vartheta) &= N(\mu_s, \sigma_s)  \\
  \mu_s &= \vartheta_{t-1} + s \\
  \sigma_s &= \xi_0 1
\end{align}
This likelihood is the probability of observing $s$ given the previous estimate
of the parameters of the force field $\phi_{t-1}$, and given the dynamics of
the system (see below).

>>>>

\documentclass[a4paper,doc,floatsintext,natbib]{apa6}
% \documentclass{article, a4paper}
\usepackage[font=small]{caption}
\usepackage{lscape}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{authblk}
\usepackage[utf8]{inputenc}
\usepackage{nameref}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{soul}
\usepackage{todonotes}

@@

\abstract{Abstract}
Humans have been shown to adapt their movements when a sudden change to the dynamics of the environment is introduced, a phenomenon called motor adaptation. If the change is reverted, the adaptation is also quickly reverted. Human are also able to adapt to multiple changes in dynamics presented separately, and to be able to switch between adapted movements on the fly. Such switching relies on contextual information which is often noisy or misleading, which affects the switch between adaptations. In this work, we use a model to explain the behavioral phenomena effected by imperfect contextual information. Specifically, we present a hierarchical computational model for motor adaptation based on exact Bayesian inference. This model explicitly takes into account contextual information and how the dynamics of context inference affect adaptation and action selection. We show how the proposed model provides a unifying explanation for four different experimentally established phenomena: (i) effects of sensory cues and proprioceptive information on switching between tasks, (ii) the effects of previously-learned adaptations on switching between tasks, in addition to  (iii) the well-studied savings, de-adaptation and spontaneous recovery.

@@

By introducing blocks of trials in which a manipulation is present (e.g. a mechanical arm exerts a force), experimenters are able to observe the dynamics of adaptation through the lens of motor error. Across many different motor adaptation experiments, now well-established phenomena have been observed, for example: (i) savings (the ability to recall previously-learned skills), (ii) quick de-adaptation (the ability to return to unmodified dynamics), (iii) anterograde interference (the interference in motor learning between opposing manipulations in dynamics) and (iv) spontaneous recovery (during trials where errors are forced to be zero, the display of behavior consistent with the previously-learned dynamics) \citep[e.g.][]{Davidson_Scaling_2004}.

@@

In this work, we show that these unexplained phenomena, alongside savings, de-adaptation, anterograde interference and spontaneous recovery, can be explained in terms of the concept of context, an idea that has recently been gaining traction in different fields \cite[e.g.][]{Sanders_Hippocampal_2020,Hunter_Contextsensitive_2021}. Intuitively, in mechanical arm experiments, one can think of one context in which the arm exerts no force and one in which it does. By inferring the current context, a participant can know how to counteract the mechanical arm and succeed in a reaching movement. In some experiments, the context is clearly inferable via sensory cues \cite[e.g.][]{Gandolfo_Motor_1996,Shadmehr_Adaptive_1994,Ethier_Spontaneous_2008}. In many other experiments, the context is not directly observable, and context inference is probabilistic and takes the form of an evidence-accumulating process across multiple trials. In this work, we propose that the behavioral consequences of this slow process of evidence accumulation can be observed in behavior and can be used to explain the phenomena outlined above.

@@

Central to the model is the concept of context. In our model, contexts are defined in terms of the task to be performed (see \fref{fig:model}A), the variables of the environment that are relevant to perform the task, the forward models used by the decision-making agent to perform the task, and the update mechanisms necessary to adapt these forward models to the variables of the environment. Together, these elements allow the agent to make predictions on future observations when this context is active, and these predictions are used to infer the context. For example, when lifting an object of unknown weight, an agent might have learned one context for heavy objects and one for light objects. When observing an object to be lifted, the agent can use its size and texture to estimate the weight of the object, which in turn allows the agent to infer the appropriate context and, with it, decide how to lift the object. These contextual cues (size and texture) do not necessarily identify the context with certainty (e.g. it could be a movie prop that looks like a rock) and integration over time might be necessary. All the variables of the model are fully described in the Methods section.

@@

It is widely accepted that in order to adapt motor commands to a novel environment, we create and update internal forward models of the outcomes of control signals \citep{Wolpert_Multiple_1998}. Here, we use this principle and chose to make use of an exact Bayesian learner. As has been discussed before \citep{Oh_Minimizing_2019}, a Bayesian learner has the advantage of not only fitting experimental data on motor adaptation, but also eliminating the need for multiple explicit time scales of learning as the learner can deal with adaptation after any number of trials. As we will show below this feature will be useful to explain specific key phenomena parsimoniously.

To describe this component of the model, we make use of the example of reaching movements. We define a forward model as follows:
\begin{equation}
p(\vartheta | s, c; \phi) = f(s, c; \phi) \label{eqn:forward-model}
\end{equation}
where $\vartheta$ are the predicted outcomes of an action (e.g. the final position of the hand in a reaching movement), $s$ is the current state (the current position of the hand), $c$ is a motor command and $\phi$ are the parameters of the forward model $f(\cdot)$. We assume that during learning the parameters $\phi$ have been fine tuned to produce an accurate forward model $f$ to guide our movements and it is these parameters which are updated during motor adaptation.

@@

\subsubsection{From observation to action selection}
Each trial starts with a new observation $\vartheta_t$, which comprises contextual sensory information (when available), as well as the outcome of the previously-selected action.

@@

\subsection{Experimental results}
In this section, we present experimentally-observed phenomena in four sections, and show that context inference, as done in our model, provides a unifying explanation for all of them. In the first section, we discuss switches between contexts, and how slow context inference affects these switches. In the second section, we focus on the effects of volatility in the environment on motor adaptation. In the third section, we discuss the effects of slow context inference on action selection. Finally, in the fourth section we discuss context inference during so-called error-clamp trials, and its effect on behavior. For each of the four sections, we selected one or two studies which are representative of the phenomenon being discussed; a more extensive review of the literature can be found in the Discussion.

@@

\subsubsection{Savings and slow/fast switching}
The term 'savings' refer to the ability of humans and other animals to remember a previously-learned adaptation and apply it without having to re-learn it. Savings is almost universally observed in human participants \citep{Brashers-Krug_Consolidation_1996,Shadmehr_Functional_1997,Medina_Mechanism_2001,Smith_Interacting_2006,Zarahn_Explaining_2008}. In an O-A-O-A experiment, for example, savings would express themselves in the second A block in the form of a much higher adaptation rate higher than that observed during the first A block.

@@

To show how adaptation changes depending on the amount of contextual information, we selected three representative experiments from two studies: \cite{Kim_Neural_2015} and \cite{Oh_Minimizing_2019} which differ in the amount of contextual information. \cite{Kim_Neural_2015} performed a cued-context visuomotor rotation experiment with three contexts with different rotation: no rotation, 40 degrees and -40 degrees. Participants performed shooting movements in blocks of trials with the same rotation. Importantly, the identity of the current context was displayed as a visual cue (colored light), making this a cued-context experiment. Consistent with our model, the authors found that switching was immediate and accurate. In \fref{fig:oh-2019}A we show the results of simulations with our model using the parameters of the task, as well as the experimental results from \cite{Kim_Neural_2015}. The correspondence between simulations and experimental results can be seen in the switches between contexts, i.e. when the solid black line (representing the true context) switches between 40, 0 and -40; in both panels (left and right), the blue line, representing the agent's adaptation, quickly follows the switches.

@@

\cite{Oh_Minimizing_2019} performed partially-cued experiments similar to \cite{Kim_Neural_2015}. The authors performed two experiments that differed in the size of the adaptation: in the first experiment, the angle of rotation was 20 degrees, while in the second experiment it was only 10 degrees. The results of their experiments can be seen in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, respectively, alongside simulations with our model. Participants in the first experiment (\fref{fig:oh-2019}B) first learned the adaptation in $A_1$. From the $A_1 - O_2$ transition onward, participants showed immediate switching (with a one-trial lag) between A and O (both ways), which can be seen in terms of the blue line (adaptation) closely following the switches in the black line (true rotation). In the second experiment (\fref{fig:oh-2019}C), switching happens more slowly, with adaptation (blue line) lagging behind the switches in the real context (black line), and slowly catching up. As can be seen in the left panels in \fref{fig:oh-2019}B and \ref{fig:oh-2019}C, the same model parametrization produces fast, accurate switches when the adaptation is large (B), and slow, noisy switches when it is low (C). This difference is explained by our model in terms of the size of the adaptation: as the adaptation is smaller (10 degrees), it becomes more difficult to distinguish errors made by incorrectly inferring the context from the noise due to trial-to-trial variation in motor output. Because of this, the model requires more evidence (i.e. more trials) to infer a switch in contexts.

@@

\subsubsection{Uncertainty over contexts affects action selection}
As with learning, our model selects actions (motor commands) based on context inference. If the identity of the current context is known, the forward model for this context is used to select the current action. However, if uncertainty over the context exists, the selected action is influenced by all the possible contexts, with a weight directly related to how likely each one of those contexts is (see \eref{eq:action-selection}).

@@

\subsubsection{Action selection in error-clamp blocks}
During error-clamp blocks at the end of block sequences, participants' behavior can be divided in two phases: (1) Spontaneous recovery, during which participants' behavior is consistent with a previously-encountered context (e.g. in an O-A-E experiment, consistent with A); this phase, when present, is seen during the early trials of the E block. (2) A slow return to baseline (i.e. no observable adaptation, consistent with the baseline context), which can last as long as hundreds of trials \citep{Brennan_Decay_2015}. However, the direction of the spontaneous recovery, its duration, the delay before it is observed, the speed of the return to baseline and the final asymptote of the adaptation seem to vary greatly depending on the experiment \citep{Brennan_Decay_2015,Vaswani_Decay_2013,Smith_Interacting_2006,Shmuelof_Overcoming_2012}.

@@

Our model also reproduces the variability in the lag before adaptation starts to drop to zero after the E block starts. In the model, this lag is due to a period in which context inference has not ``caught on'' to the change of context; during this period, behavior remains consistent with that of the previously-observed block, as can be seen in the experimental data as well. To show this, we show in \fref{fig:vaswani-2013}D the drop in the inferred probability of the previous context for each simulated run (one line per run, 10 runs per group, all groups). The start and speed of the drop depend on the run (i.e. the participant), reproducing the different lags observed in experimental data. This can also be seen directly in context inference, displayed as the shaded areas in \fref{fig:vaswani-2013}C.

@@

\subsubsection{Volatility has no effect on learning}
Experiments \citep{Herzfeld_memory_2014} have shown an effect of the volatility of the environment (i.e. unpredictable switching between contexts) on measured learning rate. Our model predicts that this effect would be greatly reduced if reliable contextual cues were introduced: if a participant can infer the context of the current trial before any decision or observation has occurred, the learning rate would not be affected by the volatility of the environment. In the experiments by \cite{Herzfeld_memory_2014}, participants had to adapt to two contexts, and the context could change from trial to trial, with a fix probability. This probability was different for two groups of participants. Our model predicts that no discernible differences would be observed between the participants that performed the task in a highly volatile environment, and those that performed in a very stable environment.

@@

\subsubsection{Multi-source integration}
The model also predicts an effect reminiscent of multisensory integration \citep[e.g.][]{Ernst_Humans_2002}: in order to integrate contextual information from conflicting sources (e.g. probabilistic visual cues and noisy endpoint feedback), the weight placed on a source increases with its reliability (e.g. the inverse of its variance). Such integration would manifest itself in experiments in which observations are noisy, as in the experiments by \cite{Kording_Bayesian_2004}, in which the position of the finger was obscured and instead participants are shown a blurry cursor which was some times shifted from its real position. If the added observation noise gives evidence for a particular context (the true underlying context or another one) and a visual cue gave partial information for another context, the participants' behavior would be more consistent with the most reliable source of contextual information.

@@

Before proceeding with the model description, let us introduce nomenclature:
\begin{itemize}
\item $t$: Discrete time at the moment an observation is made, before inference is done. 
Here taken as the trial number.
\item $\vartheta_t$: Observation. Note that this is assumed to be a noisy
observation of the generalized state (see below)
\item $s_t$: Hidden state, e.g. the position of the hand in reaching
experiments. It does not depend on the context.
\item $c_t$: Motor command (action) taken after observing $\vartheta_t$ and inferring the context.
\item $\zeta_t$: Categorical variable representing the context inferred after observing $\vartheta_t$.
\item $\omega_t$: Contextual information. This term includes, e.g. visual cues,
proprioceptive feedback, performance feedback.
\item $z_t$: Generalized state, $z_t = \{s_t, \zeta_t\}$. Observed noisily
as $\vartheta_t$.
\item $m_i$: Forward model for context $\zeta_i$.
\item $\phi_i$: Parametrization of the forward model $m_i$. In adaptation experiments, this
can be interpreted as the inferred size of the adaptation.
\end{itemize}

Context inference can work through context-specific (possibly unreliable) sensory cues,
or through the dynamics of the system (e.g. expected outputs of motor
commands), using prediction error as the force behind inference. In this case,
the generative model can be written as follows:
\[
p(z_t, \vartheta_t | z_{t-1}, c_{t-1}) = p(\vartheta_t | z_t)p(z_t|\z_{t-1}, c_{t-1})
\]
and inferring the current (hidden) generalized state is done with Bayes' theorem:
\[
  q(z_t | \vartheta_t, \vartheta_{t-1}, c_{t-1}) \propto p(\vartheta_t | z_t)p(z_t|\vartheta_{t-1}, c_{t-1})
\]

Additionally, after having inferred the current state, motor adaptation is given by
\[
q(\phi | z_t, z_{t-1}, c_{t-1}) \propto p(z_t | z_{t-1}, c_{t-1}, \phi)q(z_{t-1}|\ldots)p(\phi)
\]
where $\phi$ are the parameters of the internal model for the dynamics, also called the forward model,
i.e. the function $f: (z_t, c_t) \rightarrow z_{t+1}$, which determines the beliefs of
the agent regarding how the system evolves after having taken action $c_t$. $q(z_{t-1}|\ldots)$ is the
posterior probability calculated in the previous trial.

Having inferred the generalized state (including context), a motor command is
sampled from a distribution:
\begin{equation}
p(c_t | s_t, \zeta_t) = \displaystyle\sum_{i}q(\zeta_i)p(c_t | s_t, \zeta_i)
\end{equation}
where $i$ runs over all contexts and $p(a_t | ... m)$ is the probability distribution over actions
prescribed by the forward model $m_i$ for context $\zeta_i$.

@@

\subsubsection{Mathematical implementation}
The generalized state is given by:
\begin{equation}
z_t = \{s_t, \zeta_t\}
\end{equation}
where $s_t$ is, e.g. in a reaching task, the current position of the hand in
Cartesian coordinates, where the origin is the starting point. We assume that
motor commands are issued every $\Delta t$ for simplicity. At the beginning of
each time interval, the context is inferred combining the contextual
information and the prediction error of the outcome of the previous motor
command. This yields:
\begin{align}
  \begin{split}
  q(\zeta_t) &= q(\zeta_t | \omega_t)q(\zeta_t | s_t, s_{t-1}, c_{t-1}) \\ \label{eqn:estimated-context}
  q(\zeta_t | \omega_t) &\propto p(\omega_t | \zeta_t)p(\zeta_t) \\
  q(\zeta_t | s_t, s_{t-1}, c_{t-1}) &\propto p(s_t | \zeta_t, s_{t-1}, c_{t-1})p(\zeta_t)
  \end{split}
\end{align}
The terms $p(\zeta_t)$ refer to the prior probability of the context at the
beginning of the trial: at the first trial, this refers to prior beliefs over
which contexts are more common/likely, which we call $p_0(\zeta_t)$. At each
subsequent trial, it incorporates the belief that has so far been accumulated,
given previous trials. Note that certain events, such as the start of a new
block of trials (see main text) can return an agent to a state in which
$p_0(\zeta_t)$ becomes relevant.

We do not directly model hand movements or eye position. Instead, we model
systems in the motor error space: the position $s_t = 0$ represents a trial in
which no motor error occurred. We make use of the fact that in most
experiments, motor adaptation needs to happen in one of two directions
(e.g. clockwise vs. counter-clockwise) to keep the error space
one-dimensional. With this, $s_t > 0$ represents motor error in one direction
(e.g. clockwise) and $s_t < 0$, error in the other direction. We made this
choice because in this work we modeled different experiments with
different state spaces, but with the commonality of (at most)
two-dimensional error spaces. In order to model a specific experiment in its
own physical space (e.g. hand positions in a reaching task), only the dynamical
system  $p(s_t | \zeta_t, s_{t-1}, c_{t-1})$ would need to be adapted.

We assume that the decision-making agent has a baseline internal forward model $m_O$ that
maps motor commands to outcomes:
\begin{equation}
p(s_{t+1} | s_t, c_t, m=m_O) = f_O(s_t, c_t)
\end{equation}
where $c_t$ are motor commands (actions). We further assume that participants
have learned $f_O$ throughout their lives and use it to decide which motor
command to issue under normal circumstances (i.e. no adaptation). When a
perturbation is introduced, to which the participant must adapt, we assume they
learn a new forward model $f_m^{\phi}$, where $\phi$ represents the parameters
of the function that participants must learn. For example, in a saccade task,
$\phi$ represents the new gain. Note that $\phi$ depends on the context and
participants may need to use and update multiple forward adapted models
$f_m^{\phi}$ throughout one experiment.

@@

To complete the action-selection picture, we return to context inference. As
discussed above, action selection is made via a weighted sum of the motor commands
produced by each forward model:
\begin{equation}
p(c_t | ...) \propto \displaystyle \sum_{i} q(\zeta_i)p(c_t | \zeta_i, ...)
\end{equation}
From this distribution, an action is sampled at each trial.

@@

In the following equations, we drop the dependence on the context and
observations for notational simplicity:
\begin{equation}
q(\phi | s_{1:t}) \propto p(s_{1:t} | \phi)p(\phi)
\end{equation}
where $s_{1:t}$ are the observed states up until time $t$. The likelihood is a
Gaussian:
\begin{align}
  p(s_t | \phi) &= N(\mu_s, \sigma_s)  \\
  \mu_s &= \phi + s_t \\
  \sigma_s &= \xi_0 1
\end{align}
This likelihood is the probability of observing $s$ given the previous estimate
of the parameters of the forward model, $\phi_{t-1}$, and given the dynamics of
the system (see below).
